### 2009-11-22 docs/marcello-dissertation/thesis/chapter1-introduction.tex
Review of chapter 1 after Prof. Puder's feedback.
 - More abstract introduction, without mentioning keywords
   used in the work such as NetBEAMS, DSP, KVP, etc;

 - Reduced structure

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@655 559edde2-a4d4-44e0-96fc-4035e40d7396

Sensor networks are commonly used in different areas for different purposes
such as scientific measurements and general population use \cite{sn-intro01}
\cite{sn-intro02}. In the scientific community, they serve as tools to monitor
the state of the environment by storing samples of data in specific periods of
time. For instance, the NASA Jet Propulsion Laboratory uses the Volcano
SensorWeb \cite{sn-ex02}, a sensor network which aims to collect data from
specific volcanoes in Alaska and South Pacific to be used with heuristics
regarding dangerous activities. Similarly, the National Data Buoy Center, a
department of the National Oceanic and Atmospheric Administration (NOAA)
\cite{sn-ex03}, provides online data from different buoys in different coast
shores around the world regarding water quality, temperature, tides, etc, what
characterizes sensor network for environmental monitoring \cite{sn-ex01}.
Therefore, sensor networks can be seen as a very important scientific research
and commercial field with direct application and use of its produced data.

In order to provide access to the collected sensor data, different approaches
can be taken to interrogate sensor devices indirectly by accessing its data in
places called network sink. These data nodes can be located in the network or 
in a centralized host. In this way, the query process can take place on the
network nodes themselves or on the centralized data sink. The former is
generally used when the data is collected at the sensor node on its local
storage, being shown by \cite{sn-storage01} \cite{sn-storage03}
\cite{sn-storage04} that this strategy helps mitigating energy consumption.
However, this strategy is used when data archival is not the primary reason,
given the limited resources on the sensor devices. For this reason, the the
latter strategy has been used when the collected data is primarily archived in
a centralized database system for later reuse \cite{sn-storage02}.

When it comes to data model implementation, the relational data model
\cite{relational-model} is one of the most used approaches to represent data in
sensor network databases in any of the query processing strategies, and in most cases 
with a modified version of the SQL language \cite{sn-db-newop}. However, the
design of the database system must be done with prior knowledge of the sensor
types when choosing the relational data model, as the database design must be
normalized \cite{db-normalization} for the entities already identified for the
system. For instance, the introduction of a new sensor device might represent
a potential change on the database structure by seeking a new normalized
version of the database structure. In addition to challenges of maintaining
the database schema normalized, the use of Data Provenance approaches in
sensor network data \cite{sn-provenance} usually addresses the problem related
with the lineage of data  and its full history. However, once the data model
is chosen and implemented, one of the most important non-functional
requirements in sensor networks is scalability of the persistence storage.
Since an operational sensor network can produce huge amounts of data, it must
be able to cope with the increase and data load during data gathering, and yet
provide users the same performance during data retrieval by using different
strategies such a data-centric storage \cite{sn-storage03}.

In summary, this dissertation discusses the solution of a
centralized data-centric persistence storage component for the NetBEAMS
\cite{netbeams2009}, an implementation of a sensor network that primarily
focused on environmental data. Different challenges can be related to the
selection of a technology that handles data persistence in the context of sensor
networks: the network infrastructure, the nature and location of the collected
data, the chosen data model, among others. First, NetBEAMS needs to provide
means to safely represent sensor devices' properties without the need of
restructuring any existing data model, taking into account the use cases from
the main users. For example, a marine biologist may query the the persistence
storage regarding the temperature of a specific geographic region at the San
Francisco Bay at a given time frame. Second, since a sensor network may
experiment grown in terms of number of sensors, and consequently the number of
data produced, the persistence  storage must be able to reasonably scale and
be well structured in terms of storage space with the least service
interruption. For instance, if the persistence storage is close to reach its
storage capacity, what should the network administrator do? Last, but not
least, the means to access the data for reuse is the most important features
of the system, taking into account the different data formats used by the main
users of the system. How to export specific collected data to a spreadsheet or
archive files to be published or shared with other users? For these, and many
other reasons, the persistence storage problem addressed by this work can be
seen as a very challenging experience to solve.

In view of the fact that the persistence storage of a sensor network imposes
many different challenges, and taking into account the case study NetBEAMS, this
work's main strategy focused analyzing the different specifications of data
persistence in the literature in order to find new types of mechanisms to
provide data persistence for sensor network data. In this way, the kernel of
this dissertation is focused on providing solutions to the real-world problem
face by the Romberg Tiburon Center and its use of NetBEAMS. Furthermore,
the evaluation of different technologies is discussed in order to provide a
loosely-coupled component-based solution based on the infrastructure of the 
DSP Platform. For the reason of providing a 'lighter' data model, a subtype of
the Key-Value Pair \cite{db-kvp} data model was chosen to easily capture the
description of any sensor device's properties, along with the use of time
series data models \cite{sn-provenance} for time snapshots. In that way, data
modeling process does not represent a new refactoring each time a new sensor 
type is added or removed from the persistence system, giving users means to 
retrieve data accordingly. Similarly, in contemplation of scalability and the
use of Data-Centric approach \cite{sn-storage03}, this work proposes the use of
Database Sharding \cite{db-shard01} \cite{db-shard02}, on a centralized
database system.

In order to provide such mechanism, this work first takes a look at the related
literature to study the properties of sensor networks, in the context of its
produced data and infrastructure for storage and query processing. Then, a
summarized list of taxonomies is proposed to be the basis of an empirical  the
use of shard keys based on the geographic origin of the collected data as cited to be one of the options when working with Provenance data in sensor networks \cite{sn-provenance}. As a result, the data persistence system component was conceived using the plug-and-play capability from the DSP platform without doing  any changes to the current DSP Platform, providing an scalable solution that can adjust according to any data load increase. Finally, the selected technology provides different data access
mechanisms, which directly addressed problems related to data access and
Information Fusion \cite{sn-info-fusion}, providing export-import capabilities
and visualization through the availability of APIs written in different
popular programming languages.

This dissertation report is organized as follows: the next chapter is an 
overview of the state of art on topics that supports the implementation of the
solution of this work, providing the necessary background foundation for the 
following chapters. Chapter 3 presents the design and architecture of the 
conceived persistence module, describing the requirements needed to be
addressed,  while Chapter 4 describes the implementation based on the design. 
The analysis and investigation of the implemented solution is shown through 
the analyzes of the experiment results on Chapter 5. Last, but not least, 
Chapter 6 presents the conclusions of this work, followed by the possible 
future works and recommendations presented on Chapter 7.

>>>>

Sensor networks are commonly used in different areas for different purposes
such as scientific measurements and general population use \cite{sn-intro01}
\cite{sn-intro02}. In the scientific community, they serve as tools to monitor
the state of the environment by storing samples of data in specific periods of
time. For instance, the NASA Jet Propulsion Laboratory uses the Volcano
SensorWeb \cite{sn-ex02}, a sensor network which aims to collect data from
specific volcanoes in Alaska and South Pacific to be used with heuristics
regarding dangerous activities. Similarly, the National Data Buoy Center, a
department of the National Oceanic and Atmospheric Administration (NOAA)
\cite{sn-ex03}, provides online data from different buoys in different coast
shores around the world regarding water quality, tide information, etc, what
characterizes sensor network for environmental monitoring \cite{sn-ex01}.
Therefore, sensor networks can be seen as a very important scientific research
and commercial field with direct application and use of its produced data.

In order to provide access to the collected sensor data, different approaches
can be taken to interrogate sensor devices indirectly by accessing its data in
places called network sink. Similarly, the data model and database system
location may also influence in the process of reusing the collected data. In
this way, the lack of a persistence layer for existing sensor networks requires
a deep understanding about its infrastructure, the nature and location of the
collected data, the chosen data model representation, among others. For
instance, how can a data persistence layer safely represent the collected sensor
devices' properties without the need of restructuring any existing data model?
Should the end users' skills in database system be taken into account or the
use of programming languages to access the data \cite{sn-programming-language}?
This problem is currently faced by the departent of Biology at San
Francisco State University with a marine sensor network managed by
researchers without Computer Science skills. While the data access may impose
the difficulties, technical issues may also rise while the execution of a
sensor network such as growth in terms of number of sensor devices and,
consequently, the number of data produced. Last, but not least, the means to
access the data for reuse is the most important features of the sensor
network, taking into account the different data formats used by the main users
of the system. How can specific collected sensor data be exported to other
formats such as spreadsheets or archived files meant to be shared among
research peers? For these, and many other reasons, the inception of a
persistence storage can be a very challenging problem to solve.

In view of the fact that the persistence storage of a sensor network imposes
many different challenges, this work proposes the review of the current state of
the art for sensor networks in order to provide a data persistence layer for an
existing sensor network. As a matter of fact, this work's first contribution is
the inception of a set of basic taxonomies for data persistence in sensor
networks, proposed to drive an empirical evaluation of different technologies
used in the sensor network community. Furthermore, it also takes a look at
different alternatives not yet explored through experiments for a selected
sensor network as a case study. Similarly, this work makes another novice
contribution by suggesting a ``lighter" data model and database infrastructure
that easily captures sensor network device's observations, taking into account
the non-predictable scalability of its infrastructure, as well as the aspects of
the data's nature such as identity, time and location of the collected data.

This dissertation is organized as follows: the next chapter describes the
literature review as the state of art on sensor networks and data persistence, 
providing the needed foundation for the design of taxonomies in data
persistence for sensor networks in chapter 3. Then, chapter 4 examines the
selected case study and its requirements for a persistence layer, whose
technology properties are evaluated in chapter 5 against the developed
taxonomies. The culmination of this work is chapter 6, which details the design
and implementation of a data persistence solution for the selected case
study. Likewise, chapter 7 discusses the solution by presenting the
measurements and conducted experiments, and describing the findings
and contributions of this work. Last, but not least, chapter 8 presents the
conclusions of this work, together with the possible future works not addressed
here.
